Political Polarization, Public Trust, and Healthcare Disparities: Unraveling COVID-19's Impact and Future Preparedness

### 2024-02-13 outputs/paper/paper.qmd
Intro, abstract, and some edits

---
title: "How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?"
author: 
  - Adrian Ly
  - Hannah Yu
thanks: "Code and data are available at: https://github.com/hannahyu07/US-Covid-Analysis.git"
date: today
date-format: long
abstract: "In this research, we explored the complex interplay between local health infrastructure, socio-political factors, and COVID-19 outcomes across various U.S. states and counties to understand disparities in the pandemic's impact. Our findings reveal that political polarization and healthcare access disparities significantly influenced adherence to health guidelines and the effectiveness of the public health response. Additionally, social vulnerabilities and inconsistent public health messaging further exacerbated the pandemic's impact on marginalized communities. This study underscores the critical need for targeted public health strategies that address socio-political and infrastructural disparities to enhance pandemic preparedness and response, emphasizing the importance of equity and effective communication in safeguarding public health."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Later on, this paper delves into the intricate dynamics between local health systems, socio-political influences, and their collective impact on the disparities observed in COVID-19 outcomes across the United States. By analyzing data from various states and counties, we aim to identify how differences in health infrastructure, political affiliations, and social vulnerabilities contributed to the varied effectiveness of the response to the pandemic. This investigation seeks to fill the gap in understanding the role of these factors in shaping public health outcomes during a pandemic, a critical area of inquiry given the unprecedented scale and impact of COVID-19.

Our research uncovered that political polarization significantly affected adherence to health guidelines, with stark differences in compliance and outcomes between regions with varying political leanings. Furthermore, disparities in healthcare access and the prevalence of social vulnerabilities were closely correlated with the severity of the pandemic's impact, highlighting the exacerbating effect of pre-existing inequalities. These findings underscore the necessity of developing public health strategies that are sensitive to socio-political dynamics and infrastructural disparities, to enhance resilience against future health crises.

This paper is structured to first provide a comprehensive overview of the broader context and motivations behind this study, followed by a detailed examination of the methodologies employed to investigate the research question. The results section presents the key findings, which are then discussed in the context of their implications for public health preparedness and response strategies. The conclusion summarizes the study's contributions to our understanding of the complex factors influencing pandemic outcomes and outlines recommendations for future research and policy development.

@@

Various studies have confirmed the existence of correlation between a state's political affiliation and its handling of COVID issues (@datz_2022). To testify to the claim, we collected the voting data for each of the 50 states for the 2020 presidential election. We then selected the ten states with the highest proportions of Republican votes. These voting patterns are visualized in Figure @fig-Vote, where red symbols represent Republican votes and blue symbols represent Democratic votes. Among the top ten states, Tennessee recorded the highest total number of votes, while Wyoming boasted the highest proportion of Republican votes.

@@

Following our previous analysis regarding individuals' political affiliations, we have developed Figure @fig-deaths-rates-votes, which encompasses all 50 states of the US along with their political leanings based on which party garnered the majority votes. This information is juxtaposed against their respective COVID-19 death rates.

>>>>

---
title: "Political Polarization, Public Trust, and Healthcare Disparities: Unraveling COVID-19's Impact and Future Preparedness"
author: 
  - Adrian Ly
  - Hannah Yu
  - Sakhil Goel
thanks: "Code and data are available at: https://github.com/hannahyu07/US-Covid-Analysis.git"
date: today
date-format: long
abstract: "The COVID-19 pandemic has exposed deep disparities in health outcomes across different states and counties in the United States. We use data from the American Economic Association to conduct a comprehensive analysis of how variances in local health infrastructure and socio-political factors within the United States contributed to disparities in COVID-19 outcomes. Our analysis reveals that political polarization significantly influenced adherence to health guidelines, with Republican-leaning states demonstrating slightly higher clustering around elevated death rates. Moreover, social vulnerabilities, including disparities in healthcare access, housing conditions, and ethnic demographics, emerged as key contributors to substantial variations in disease outcomes. These results highlight the urgent requirement for customized public health readiness and response tactics, focusing on precise interventions to tackle deficiencies in community health infrastructure and socio-political factors."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

@@

Various studies have confirmed the existence of correlation between a state's political affiliation and its handling of COVID issues (@datz_2022). To testify to the claim, we collected the voting data for each of the 50 states for the 2020 presidential election. We then selected the ten states with the highest proportions of Republican votes. These voting patterns are visualized in @fig-Vote, where red symbols represent Republican votes and blue symbols represent Democratic votes. Among the top ten states, Tennessee recorded the highest total number of votes, while Wyoming boasted the highest proportion of Republican votes.

@@

Following our previous analysis regarding individuals' political affiliations, we have developed @fig-deaths-rates-votes, which encompasses all 50 states of the US along with their political leanings based on which party garnered the majority votes. This information is juxtaposed against their respective COVID-19 death rates.
